---
title: Color
format: gfm
---

## Arts & crafts

For the first section of this lab, you'll do some arts & crafts in class with paper. For each exercise, you can  take pictures of your work for your own reference, but you don't need to upload anything. Just jot down some notes here. It's more for you to have a tactile experience of working with color.

Do these in any order; we have limited sets of materials so we'll rotate.

### Sorting by hue 

There are 3 sets of colors to sort by hue: one of blues, one of pastel colors in small pieces, and one of the same pastel colors in larger pieces. For each of these, lay the papers out touching side by side in a random order. Then arrange them by hue (named color, e.g. yellow, yellow-green, teal), swapping them in pairs. For example, if you're alphabetizing this set of letters:

B D A F E C 

Your first step could be swapping B & A to get A into the first position:

A D B F E C

Then swap D & B to get B into second:

A B D F E C 

and so on.

Try each set on at least 2 backgrounds. Take note of what difficulties you have, how you see colors shift depending on their neighbors, and which background is easiest.

### Qualitative colors

There are 2 types of shapes in the qualitative sets; start with the bars, not the lines. You're designing charts for a client, and these colors fit their branding palette, so you're stuck with them. You have a chart with 5 groups, so pick 5 colors. 

* Chart 1: Lay out the bar pieces touching side by side to simulate bar charts. If two neighboring bars are too similar, shuffle them around until they're more distinct. 
* Chart 2: Move the bars apart slightly so there's a gap in between them. Does this make any pairs of colors easier to tell apart?
* Chart 3: Move the pieces around to simulate stacked bars. Again, shuffle pieces if you need to to keep them distinct. 
* Chart 4: Using the skinny strips in the same 5 colors you chose, simulate a line chart. Do those colors still hold up with smaller markings?

## Code exercises

### Picking a palette

Visit [ColorBrewer](https://colorbrewer2.org) and [Carto Colors](https://carto.com/carto-colors/). For each chart below, pick a ColorBrewer or Carto palette of an appropriate type. Then apply that palette to the chart with `scale_fill_brewer` or `rcartocolor::scale_fill_carto_d`, depending on the source (and change "fill" to "color" depending on the geom type). Each function will take the palette name as its `palette` argument.

For each chart, then decide whether the default version of the palette works well. Are colors distinct enough from each other? Did you use the right type of palette (sequential, qualitative, or diverging) for the data? If the palette has a gray color, is it used in a way that makes sense? If it doesn't have a gray color, do you think it would be good to add one?

Based on those assessments, adjust your selection of colors. Get the palette as a vector (use `RColorBrewer::brewer.pal` or `rcartocolor::carto_pal`) so you can change the order of the colors however you want. Once you have your custom version of the palette, use it with `scale_fill_manual(values = new_custom_palette)` (or `scale_color_manual`).

```{r}
#| label: setup
library(ggplot2)

theme_set(theme_minimal())
locations <- c(
    "United States",
    "Maryland",
    "Baltimore city",
    "Baltimore County",
    "Anne Arundel County",
    "Howard County"
)
acs_county <- justviz::acs |>
    dplyr::filter(name %in% locations)
```

```{r}
#| label: dodged
cost_burden <- acs_county |>
    dplyr::select(level, name, total_cost_burden, total_severe_cost_burden) |>
    tidyr::pivot_longer(
        -level:-name,
        names_to = "group",
        values_to = "share"
    ) |>
    dplyr::mutate(
        group = stringr::str_remove(group, "^total_") |> forcats::as_factor()
    ) |>
    dplyr::mutate(name = forcats::as_factor(name))

ggplot(cost_burden, aes(x = name, y = share, fill = group)) +
    geom_col(position = position_dodge2()) +
    labs(title = "Cost burden and severe cost burden rates, 2024")
```


```{r}
#| label: stacked-1
age <- acs_county |>
    dplyr::select(level, name, ages00_17, ages18_34, ages35_64, ages65plus) |>
    tidyr::pivot_longer(
        -level:-name,
        names_to = "group",
        values_to = "share"
    ) |>
    dplyr::mutate(dplyr::across(c(name, group), forcats::as_factor))

ggplot(age, aes(x = name, y = share, fill = group)) +
    geom_col(position = position_fill(reverse = TRUE)) +
    labs(title = "Population by age group, 2024")
```


```{r}
#| label: stacked-2
race <- acs_county |>
    dplyr::select(level, name, white, black, latino, asian, other_race) |>
    tidyr::pivot_longer(
        -level:-name,
        names_to = "group",
        values_to = "share"
    ) |>
    dplyr::mutate(dplyr::across(c(name, group), forcats::as_factor))

ggplot(race, aes(x = name, y = share, fill = group)) +
    geom_col(position = position_fill(reverse = TRUE)) +
    labs(title = "Population by race/ethnicity, 2024")
```

```{r}
#| label: scatter-1
acs_tract <- justviz::acs |>
    dplyr::filter(county %in% locations) |>
    # lump counties besides Balt County into "other"
    dplyr::mutate(
        balt_county = forcats::fct_other(
            county,
            keep = "Baltimore County",
            other_level = "Nearby counties"
        )
    )

ggplot(acs_tract, aes(x = foreign_born, y = diversity_idx, color = county)) +
    geom_point(alpha = 0.7) +
    labs(title = "Diversity index vs % foreign-born by tract, 2024")
```

```{r}
#| label: scatter-2
ggplot(
    acs_tract,
    aes(x = foreign_born, y = diversity_idx, color = balt_county)
) +
    geom_point(alpha = 0.7) +
    labs(title = "Diversity index vs % foreign-born by tract, 2024")
```

### Revising a palette

I would generally caution against editing the standard, research-based palettes, especially sequential and diverging ones from ColorBrewer or Carto. But it's still good to see how you might tweak those palettes and how they change across different color spaces.

Get a copy of a premade sequential palette with only 3 colors. Print it to see its hex codes, then copy those codes and paste them as inputs in the [Chromatta tool](https://obumbratta.com/chromatta). 

```{r}
#| label: pal-1
# with colorbrewer
pal_to_adjust <- RColorBrewer::brewer.pal(n = 3, name = "") # plug in a name here
# or with carto
# pal_to_adjust <- rcartocolor::carto_pal(n = 3, name = "") # plug in a name here
print(pal_to_adjust)
```

In Chromatta, set the number of colors to 5. Make sure correct lightness and intelligent interpolation are both switched off. Start with the LAB color space selected (first button). Then switch to HSL. What do you notice about the transitions between colors and how that changes? What about after turning on correct lightness and color interpolation?

Now remove the middle color so you just have the two end points. Switch between different color spaces and settings, and check out the sample visuals panel. Once you find a 5-color palette you like, export it by clicking the JavaScript copy button (looks like `[ ]`).

JavaScript uses arrays similar to vectors in R, so you'll just need to change the syntax slightly. Paste it in the next code chunk, then change it from being wrapped in brackets to being wrapped in the `c()` function, e.g. change

```javascript
['#d1eeea', '#a1c7ca', '#74a1ac', '#4c7b90', '#2a5674']
```

to

```r
c('#d1eeea', '#a1c7ca', '#74a1ac', '#4c7b90', '#2a5674')
```

```{r}
#| label: pal-paste
pal_chrom <- c() # replace with your palette from chromatta
```

Then use that new palette with `scale_fill_manual` in the chart below.


```{r}
#| label: age-bars-redo
ggplot(age, aes(x = name, y = share, fill = group)) +
    geom_col(position = position_fill(reverse = TRUE)) +
    labs(title = "Population by age group, 2024") +
    scale_fill_manual(values = pal_chrom)
```

Compare it to your previous version of this chart (population by age). Which do you like better? What works and doesn't work with each?