---
title: Text, audience, accessibility
format: gfm
---

Get set up by sourcing the formatters utility script, plus any other scripts of helpers you've defined. Pick a qualitative color palette, such as from Carto Colors or ColorBrewer. You're now starting to do the sorts of refining tasks that should set you up for a good grade on your midterm project!

```{r}
#| label: setup
# source formatters script. add any other scripts in your utils folder you have
source(here::here("utils/formatters.R"))
source(here::here("utils/palettes.R"))
source(here::here("utils/helper_functions.R"))
library(ggplot2)
library(ggtext)
library(showtext)
library(justviz)

# placeholders
qual_pal <- rcartocolor::carto_pal(name = "Pastel")
```

## Adding text within a chart

Go back to lab 2 where we defined a function to place labels slightly inward from the end of a bar. Copy and paste that code here. 

```{r}
#| label: label-nudge-function
# placeholder---this just returns the input value
label_nudge <- function(y) {

    # Calculate an offset equal to 5% of the largest bar
    offset <- 0.05 * max(y)

    # Place labels slightly below the top of each bar
    label_position <- y - offset

    return(label_position)
}
```

Add code to the definition of dental_visit to add 2 new variables: `nudged_value`, a nudged position for the labels, and `pct_value`, where you've formatted the labels as percentages. Do this either in 1 or 2 `dplyr::mutate` calls. 

You can use the `percent` function that's in the formatters script, or define a new one to fit a different number of decimal places (`accuracy` argument), since the CDC data includes 1 decimal place.

If the label placement doesn't look the way you want, add or adjust arguments to your `label_nudge` function until you like the way things are calculated.

One helpful pair of `geom_text` arguments is `hjust`/`vjust`, which set the horizontal and vertical justifications, respectively, and range from 0 (aligned at the left/bottom) to 1 (aligned at the right/top). For vertical bars, you want the labels centered horizontally (`hjust = 0.5`); for horizontal bars, you want labels centered vertically (`vjust = 0.5`). Those are both the defaults, but I've set them here explicitly.

```{r}
#| label: label-nudge-1

# new scales func, pretty easy to setup so just did it here
pct1 <- scales::label_percent(accuracy = 0.1)

dentist_chart <- ggplot(dental_visit, aes(x = location, y = value)) +
    geom_col(fill = qual_pal[3]) +
    geom_text(
        aes(y = nudged_value, label = pct_value),
        size = 14,
        size.unit = "pt",
        fontface = "bold",
        hjust = 0.5,
        vjust = 1,
        color = "#222222"
    ) +
    labs(
        title = "Percentage of People Who Went to the Dentist in 2022",
        x = "Location",
        y = "Percentage"
    ) +
    scale_y_continuous(labels = percent100) +
    scale_x_discrete(labels = scales::label_wrap(12))

dentist_chart
```

Does that function work as well when the chart is flipped to horizontal? Adjust arguments of the function and in your `geom_text` call. 

```{r}
#| label: label-nudge-2
#| 
dentist_chart_h <- ggplot(dental_visit, aes(x = value, y = location)) +
    geom_col(fill = qual_pal[4]) +
    geom_text(
        aes(x = nudged_value, label = pct_value),
        size = 14,
        size.unit = "pt",
        fontface = "bold",
        hjust = 1,
        vjust = 0.5,
        color = "#222222"
    ) +
    labs(
        title = "Percentage of People Who Went to the Dentist in 2022",
        x = "Percentage",
        y = "Location"
    ) +
    scale_x_continuous(labels = percent100) +
    coord_cartesian(reverse = "y")

dentist_chart_h
```

## Audience

Copy over the code for one of your lab 9 charts with your nice theme applied. Imagine you're a consultant to a local government agency (pick one depending on what data is in your chart). Make 2 revisions of your chart:

- One version you'd use to present your research to a meeting of officials from the agency. You can use whatever type of title you want, but it should be at no higher than a 10th grade reading level.
- One version you'd suggest they use for a public awareness campaign on the issue, such as social media posts, handouts at clinics or government offices, posters on bus shelters, etc. Use narrative titles ("Baltimore County has a higher...") at no higher than a 7th grade reading level.

When you design for each audience, consider your chart type and your use of color, text, and annotations. Add annotations that you think will help guide your reader if necessary, but don't add clutter.

Use [wordcalc](https://www.wordcalc.com/readability/) or another online Flesch-Kincaid grade level checker. (On wordcalc, you want the score that says "Flesch-Kincaid Grade Level").

For both charts, write alt text that will explain the chart to someone using a screen reader (put it in the `fig-alt` comment). Follow the recommendations from [Amy Cesal's post](https://medium.com/nightingale/writing-alt-text-for-data-visualization-2a218ef43f81?sk=32db60d651933b5ac2c5b6507f3763b5).

```{r}
#| label: og-chart-code

chart2 <- tracts |>
  dplyr::filter(!is.na(bachelors)) |>
  dplyr::mutate(county = stringr::str_remove(county, " County")) |>
  ggplot(aes(x = bachelors,
             y = forcats::fct_reorder(county, bachelors, .fun = median))) +
  geom_boxplot(outlier.shape = NA,
               fill = NA,
               color = sunlight_pal$text_light,
               linewidth = px_lw(2),
               width = 0.5) +
  geom_point(position = position_jitter(height = 0.15),
             color = sunlight_pal$main_blue,
             alpha = 0.5,
             size = 1.1) +
  scale_x_continuous(labels = scales::label_percent(),
                     limits = c(0, NA),
                     expand = expansion(mult = c(0, 0.05))) +
  labs(
    title = paste0(
      "**Bachelor's degrees** by county in Central MD<br>",
      "<span style='font-family:georgia; font-style:italic; font-size:12px; color:",
      sunlight_pal$text_light, "'>",
      "Each dot is a census tract; boxes show the middle 50% of tracts in each county.</span>"
    ),
    x       = "% with bachelor's degree",
    y       = NULL,
    caption = "**Source:** *U.S. Census Bureau*"
  ) +
  sunlight()

chart2
```


```{r}
#| label: chart-agency
#| fig-alt: "Box plot showing a dot for each census tract representing the percent of adults with a bachelor's degree in five central Maryland counties: Howard, Montgomery, Frederick, Carroll, and Washington. Counties are sorted by their middle value, and Frederick County is highlighted in teal. The chart is meant to show how Frederick compares with nearby counties and how much rates vary between tracts inside each county. Howard County has the highest middle value, Washington County has the lowest, and Frederick sits in between."

focus <- "Frederick County" 

agency_df <- tracts |>
    dplyr::filter(!is.na(bachelors)) |>
    dplyr::mutate(
        county = forcats::fct_reorder(county, bachelors, .fun = median),
        is_focus = county == focus
    )

focus_median <- median(agency_df$bachelors[agency_df$is_focus])

chart_agency <- ggplot(agency_df, aes(x = bachelors, y = county)) +
    geom_boxplot(
        outlier.shape = NA,
        fill = NA,
        color = sunlight_pal$text_light,
        linewidth = px_lw(2),
        width = 0.5
    ) +
    geom_point(
        data = dplyr::filter(agency_df, !is_focus),
        position = position_jitter(height = 0.15, seed = 1),
        color = sunlight_pal$text_light,
        alpha = 0.4,
        size = 1.1
    ) +
    geom_point(
        data = dplyr::filter(agency_df, is_focus),
        position = position_jitter(height = 0.15, seed = 1),
        color = sunlight_pal$main_blue,
        alpha = 0.8,
        size = 1.8
    ) +
# agency prob cares abt what is in its own county, so pretending theyre from frederick
  annotate(
        "text",
        x = max(agency_df$bachelors[agency_df$is_focus]),
        y = focus,
        label = paste("Middle tract:", percent100(focus_median)),
        hjust = -0.1,
        vjust = 0.5,
        size = 4,
        fontface = "bold",
        family = main_font,
        color = sunlight_pal$main_blue
    ) +
    scale_color_manual(
        values = c("TRUE" = sunlight_pal$main_blue, "FALSE" = sunlight_pal$text_light)
    ) +
    scale_x_continuous(
        labels = percent100,
        limits = c(0, NA),
        expand = expansion(mult = c(0, 0.05))
    ) +
    guides(color = "none") +
    labs(
        title = paste0(
            "**Share of adults with a bachelor's degree, by county**<br>",
            "<span style='font-family:georgia; font-style:italic; font-size:12px; color:",
            sunlight_pal$text_light, "'>",
            "Each dot is a census tract. Boxes show the middle half of tracts in each county.</span>"
        ),
        x = "Adults with a bachelor's degree",
        y = NULL,
        caption = "**Source:** *U.S. Census Bureau*"
    ) +
    sunlight()

chart_agency
```


```{r}
#| label: chart-public
#| fig-alt: ""
#
```

## Contrast

Copy your vertical chart of dental visits (the chunk labeled "label-nudge-1"). You'll test contrast between the labels and the bars using the [APCA contrast checker](https://apcacontrast.com/). Plug in your label color (text, on the left) and your bar fill color (background, on the right). For text like this, you need a Lc value of at least 60. If you don't have enough contrast, adjust the colors until you do.

Make two versions of this chart, one with dark labels (`"#444444"` or lower) and one with light labels (`"#eeeeee"` or higher). The fill colors for these charts will have to be different.

The RStudio add-in [colourpicker](https://github.com/daattali/colourpicker) helps convert between R's named colors (e.g. gray10) and hex codes (#1a1a1a). You can also look up gray colors from [R Charts](https://r-charts.com/colors/), or use the function `col2hex` which I've added to a utility script you can source.

```{r}
#| label: contrast-labels-dark
#

```

```{r}
#| label: contrast-labels-light
#

```
