---
title: Batch generating charts
format: gfm
---

Feel free to ignore this!

By Keaton's request, here are notes on batch generating lots of charts. My usual method is reshape, split, apply: reshape the data into a long format, split into groups (variable, etc), apply a function to each group.


```{r}
#| label: setup
library(ggplot2)

acs_split <- justviz::acs |>
    dplyr::filter(level == "tract") |>
    # reshape, making indicator column a factor
    tidyr::pivot_longer(
        -level:-name,
        names_to = "indicator",
        names_ptypes = factor()
    ) |>
    # split into a list of data frames based on indicator
    split(~indicator)

# varify that this is a list, that each list item is a data frame,
# and check the length
class(acs_split)
class(acs_split[[1]])
length(acs_split)
```

A batch list of 38 histograms

```{r}
#| label: batch-charts
# map_* functions map over a list to apply a function
# imap means it takes 2 arguments, the list item & its index
# because it's split by a factor, the index is factor levels
purrr::imap(acs_split, function(df, id) {
    ggplot(df, aes(x = value)) +
        geom_histogram(bins = 50) +
        labs(title = stringr::str_glue("ACS data: {id} by tract"))
})
```

In base R, you could instead take a vector of column names for the columns you want to plot, then iterate over that with `lapply` (kind of the base equivalent of `purrr::map`). `hist` just needs a vector instead of a whole data frame, so you can subset the data frame for just the vector of interest and plot it.

```{r}
#| label: base-r
# get the names of columns other than the ones for geography
# other ways to do this
cols <- names(justviz::acs)[
    names(justviz::acs) %notin% c("level", "name", "county")
]

# alternatives:
# select columns programmatically by data type
# names(justviz::acs)[unlist(lapply(justviz::acs, is.numeric))]
# drop columns by index
# names(justviz::acs)[-1:-3]

# assign back to a variable just to avoid printing a bunch of NULLs
plots <- lapply(cols, function(col) {
    # extract column as a vector for rows where level == tract
    col_data <- justviz::acs[[col]][justviz::acs$level == "tract"]
    hist(col_data, main = col)
})
```

